Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-20 doc/asplos13/grappa.tex
quick pass on sec 4. will do another pass on 3-4 once the TODOS are done.

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.5\columnwidth]{figs/grappa-crappy}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}\TODO{should we use the canonical Grappa picture here agumented withe the software stack?}}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

In this section we describe the Grappa runtime system
(Figure~\ref{fig:grappa}). 
%%% Moved to intro
% Grappa is a software runtime system that allows a commodity distributed-memory
% HPC cluster to be programmed as if it were a single large shared-memory
% machine.
% Grappa is designed to
% smooth over some of the performance discontinuities in commodity
% hardware, giving good performance when there is little locality to be
% exploited while allowing the programmer to exploit it when it is
% available.
Grappa provides three key capabilities. The first is a distributed
shared \textbf{global memory} with high aggregate random access
bandwidth. The second is a \textbf{tasking} system with lightweight
multithreading and automatic load balancing through work
stealing. Finally, under the hood, a \textbf{communication} layer uses
network packet scheduling and aggregation to achieve high performance
while remaining largely invisible to applications. We now discuss the programmer's view of
Grappa's main capabilities, including a subset of the Grappa API. In
Section~\ref{implementation}, we discuss the details of their
implementation.

\TODO{Should we have an example of code here, and walk thru as we explain the
components and parts of the API?}

@@

The basic unit of execution in Grappa is the {\em task}. Tasks are
small; just a function pointer and arguments. Tasks are only
queued when they are spawned; later, when resources are free, they are
allocated a stack, bound to a core, and executed.

>>>>

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.5\columnwidth]{figs/grappa-crappy}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}\TODO{should we use the canonical Grappa picture here augmented withe the software stack?}}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

In this section we describe the Grappa runtime system
(Figure~\ref{fig:grappa}). 
Grappa provides three key capabilities. The first is a distributed
shared \textbf{global memory} with high aggregate random access
bandwidth. The second is a \textbf{tasking} system with lightweight
multithreading and automatic load balancing through work
stealing. Finally, under the hood, a \textbf{communication} layer uses
network packet scheduling and aggregation to achieve high performance
while remaining largely invisible to applications. We now discuss the programmer's view of
Grappa's main capabilities, including a subset of the Grappa API. In
Section~\ref{implementation}, we discuss the details of their
implementation.

\TODO{Should we have a code example here, and walk thru it as we explain the
components and parts of the API?}

@@

The basic unit of execution in Grappa is a {\em task}. Tasks are
small; just a function pointer and arguments. Tasks are only
queued when they are spawned; later, when resources are free, they are
allocated a stack, bound to a core, and executed.

### 2012-07-20 doc/asplos13/implementation.tex
quick pass on sec 4. will do another pass on 3-4 once the TODOS are done.

Grappa implements a software distributed shared memory. This memory is
exposed through two address spaces; one optimized for global access to
node-local data, and another optimized for high aggregate random
access bandwidth to low locality data.

\paragraph{2D global addresses} All global addresses in Grappa are 64-bit values. Each byte of global
memory is associated with a {\em home core}; this core is responsible
for allocating and modifying that memory. Since today's commodity
processors support a 48-bit virtual address space \TODO{cite?}, Grappa
uses the spare bits to distinguish which address space a global
address belongs to, and to represent the home core of a global
address.

The first global address space is the {\em two-dimensional} address
space. This uses a traditional PGAS addressing model, where each
address is a tuple of a rank in the job (or global process ID) and an
address in that process. The lower 48 bits of the address hold a
virtual address in the process. The top bit is set to indicate that
this a 2D address. This leaves 15 bits for network endpoint ID, which
limits our scalability to $2^15$ endpoints.

Any local data can be made accessible by other nodes in the system by
wrapping the address and node ID into a 2D global address. This
address can then be accessed with delegate or cache operations. At the
destination the address is converted into a canonical x86 address by
replacing the upper bits with the sign-extended upper bit of the
virtual address.

@@

\paragraph{Caches} Cache operations are used to exploit spatial or temporal
locality. They allow a single task to buffer a range of data, read or
modify it, and write the modified version back to its home node. All
acquires and releases are expressed explicitly by the
programmer. Cache operations may operate on data that spans multiple
blocks of the linear address space; the programmer provides the
starting address and length, and the cache issues as many active
messages as necessary to retrieve all necessary data.

>>>>

Grappa implements a software distributed shared memory. This memory is exposed
through two distinct, non-overlapping address spaces; one optimized for global
access to node-local data, and another optimized for high aggregate random
access bandwidth to low locality data.

All global addresses in Grappa are 64-bit
values. Each byte of global memory is associated with a {\em home core}; this
core is responsible for allocating and modifying that memory. Since today's
commodity processors support a 48-bit virtual address space \TODO{cite?},
Grappa uses the spare bits to distinguish which address space a global address
belongs to, and to represent the home core of a global address.

\paragraph{2D global addresses} The first global address space is the {\em
two-dimensional} address space. This uses a traditional PGAS addressing model,
where each address is a tuple of a rank in the job (or global process ID) and
an address in that process. The lower 48 bits of the address hold a virtual
address in the process. The top bit is set to indicates that the reference is
a 2D address. This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^15$ endpoints.

Any node-local data can be made accessible by other nodes in the system by
wrapping the address and node ID into a 2D global address. This address can
then be accessed with delegate and can also be cached by other nodes. At the
destination the address is converted into a canonical x86 address by replacing
the upper bits with the sign-extended upper bit of the virtual address.

@@

\paragraph{Explicit caches} Grappa cache operations are used to exploit
spatial or temporal locality. They allow a single task to buffer a range of
data, read or modify it, and write the modified version back to its home node.
All fetch and update operations of cache data are expressed explicitly by the
programmer. Cache operations may operate on data that spans multiple blocks of
the linear address space; the programmer provides the starting address and
length, and the cache issues as many active messages as necessary to retrieve
all necessary data.
